Robust Co-occurrence Quantification for Lexical Distributional Semantics

### 2016-04-26 paper.tex
Submission
Pre Student ACL

\title{Robust co-occurrence quantification for low-dimensional lexical distributional semantics}

@@

\begin{document}
\def\emnlp/{\textit{KS2013}}
\def\PhraseRel/{PhraseRel}

\def\PMI/{$1 \operatorname{PMI}$}
\def\SPMI/{$1 \operatorname{SPMI}$}
\def\CPMI/{$1 \operatorname{CPMI}$}
\def\SCPMI/{$1 \operatorname{SCPMI}$}

\def\NPMI/{$n \operatorname{PMI}$}
\def\NSPMI/{$n \operatorname{SPMI}$}
\def\NCPMI/{$n \operatorname{CPMI}$}
\def\NSCPMI/{$n \operatorname{SCPMI}$}

\def\logNPMI/{$\log n\operatorname{PMI}$}
\def\logNSPMI/{$\log n\operatorname{SPMI}$}
\def\logNCPMI/{$\log n \operatorname{CPMI}$}
\def\logNSCPMI/{$\log n \operatorname{SCPMI}$}

\maketitle
\begin{abstract}
\input{abstract.tex}
\end{abstract}

@@

\newcite{TACL570} propose optimisations for co-occrrence-based distributional models, using parameters adopted from predictive models \cite{mikolov2013efficient}: \emph{shifting} and \emph{context distribution smoothing}. Their experiments and thus their parameter  recommendations use high-dimensional vector spaces with word vector dimensionality of almost \textbf{200K}, and many recent state-of-the-art results in lexical distributional semantics have been obtained using vectors with similarly high dimensionality  \cite{baroni-dinu-kruszewski:2014:P14-1,kiela-clark:2014:CVSC,lapesa2014large}.

In contrast, much work on \emph{compositional} distributional semantics employs vectors with much fewer dimensions: e.g.~\textbf{2K} \cite{Grefenstette:2011:ESC:2145432.2145580,kartsadrqpl2014,milajevs-EtAl:2014:EMNLP2014}, \textbf{3K} \cite{Dinu:2010:MDS:1870658.1870771,milajevs-purver:2014:CVSC} or \textbf{10K} \cite{polajnar-clark:2014:EACL,Baroni2010nouns}. The most common reason thereof is that these models  assign tensors to functional words. For a vector space $V$ with $k$ dimensions,  a tensor $V \otimes V \cdots \otimes V$ of rank $n$ has $k^n$ dimensions. Adjectives and intransitive verbs have tensors of rank 2, transitive verbs are of rank 3; for coordinators, the rank can go up to 7. Taking $k = \textbf{200K}$ already results in a highly intractable tensor of $\textbf{8} \times \textbf{10}^{\textbf{15}}$ dimensions for a transitive verb. 
%
This mismatch in vector dimensionality between lexical and compositional tasks gives rise to a number of questions:
\begin{compactitem}
\item To what extent does model performance depend on vector dimensionality?
\item Do parameters influence high-dimensional and low-dimensional models similarly? Can the findings of \newcite{TACL570} be directly applied to low-dimensional models?
\item If not, can we derive suitable parameter selection heuristics which take account of dimensionality?
\end{compactitem}

@@

In addition to the extreme $1$ and $n$ weightings, we also experiment with a $\log n$ weighting.
% \textbf{MP: cite ref for this?}

@@

\newcite{TACL570} show that performance is affected by smoothing the context distribution $P(x)$:
%
\begin{equation}
  \label{eq:cds}
  P_{\alpha}(x) = \frac{n(x)^{\alpha}}{\sum_{c}n(c)^{\alpha}}
\end{equation}
%
We experiment with $\alpha$=1 (no smoothing) and $\alpha$=0.75. We call this estimation method \emph{local context probability}; we can also estimate a \emph{global context probability} based on the size of the corpus $C$:
%
\begin{equation}
  \label{eq:cds-nan}
  P(x) = \frac{n(x)}{|C|}
\end{equation}
%
%where $|C|$ is the size of the used corpus.

@@

We evaluate these heuristics by comparing the performance they give on SimLex-999 against that obtained using the best possible parameter selections (determined via an exhaustive search at each dimensionality setting). We also compare them to the best scores reported by \cite{TACL570} for their model, word2vec-SGNS \cite{mikolov2013efficient} and GloVe \cite{pennington2014glove} -- see Figure~\ref{fig:best-simlex} (only the better-performing SPMI and SCPMI are shown).

@@

Finally, to see whether the heuristics transfer robustly, we repeat this comparison on the MEN dataset (see Figures~\ref{fig:best-men},\ref{fig:global-best-men}). Again, PMI and CPMI follow the best possible setup, with SPMI and SCPMI showing only a slight drop below ideal performance; and again, the heuristic settings give performance close to the optimum, and significantly higher than average or standard parameters.

% \textbf{MP: is it worth including a line for the *worst* possible parameter selection too? Or for the average; or for the Levy/Goldberg "standard" parameters? That way the reader can see how much benefit the heuristics give. }

@@

This paper presents a systematic study of co-occurrence quantification focusing on the selection of parameters presented in \newcite{TACL570}. We replicate their recommendation for high-dimensional vector spaces, and show that with appropriate parameter selection it is possible to achieve comparable performance with low-dimensional spaces, and propose a set of model selection heuristics that maximizes performance.

In general, model performance depends on vector dimensionality (the best setup with 50K dimensions is better than the best setup with 1K dimensions by 0.03 on SimLex-999). Low-dimensional spaces benefit from being dense and unsmoothed  ($k < 1$, global context probability); while high-dimensional spaces are better sparse and smooth  ($k > 1$, $\alpha = 0.75$). However, these heuristics do not guarantee the best possible result for unweighted and $n$-weighted models because of the high variance of the corresponding scores. Based on this we suggest the use of \logNSPMI/ or \logNSCPMI/ with dimensionality of at least 20K to ensure good performance on lexical tasks.

>>>>

\title{Robust Co-occurrence Quantification for Lexical Distributional Semantics}

@@

\begin{document}

\def\PMI/{$\operatorname{1PMI}$}
\def\SPMI/{$\operatorname{1SPMI}$}
\def\CPMI/{$\operatorname{1CPMI}$}
\def\SCPMI/{$\operatorname{1SCPMI}$}

\def\NPMI/{$\operatorname{nPMI}$}
\def\NSPMI/{$\operatorname{nSPMI}$}
\def\NCPMI/{$\operatorname{nCPMI}$}
\def\NSCPMI/{$\operatorname{nSCPMI}$}

\def\logNPMI/{$\operatorname{lognPMI}$}
\def\logNSPMI/{$\operatorname{lognSPMI}$}
\def\logNCPMI/{$\operatorname{lognCPMI}$}
\def\logNSCPMI/{$\operatorname{lognSCPMI}$}

\maketitle
\begin{abstract}
Previous optimisations of parameters affecting the word-context association measure used in distributional vector space models have focused either on high-dimensional vectors with hundreds of thousands of dimensions, or dense vectors with dimensionality of few hundreds; but dimensionality of few thousands is often applied in compositional tasks as it is still computationally feasible and does not require the dimensionality reduction step. We present a systematic study of the interaction of the parameters of the association measure and vector dimensionality, and derive parameter selection heuristics that achieve performance across word similarity and relevance datasets competitive with the results previously reported in the literature achieved by highly dimensional or dense models.
\end{abstract}

@@

\newcite{TACL570} propose optimisations for co-occurrence-based distributional models, using parameters adopted from predictive models \cite{mikolov2013efficient}: \emph{shifting} and \emph{context distribution smoothing}. Their experiments and thus their parameter  recommendations use high-dimensional vector spaces with word vector dimensionality of almost \textbf{200K}, and many recent state-of-the-art results in lexical distributional semantics have been obtained using vectors with similarly high dimensionality  \cite{baroni-dinu-kruszewski:2014:P14-1,kiela-clark:2014:CVSC,lapesa2014large}.

In contrast, much work on \emph{compositional} distributional semantics employs vectors with much fewer dimensions: e.g.~\textbf{2K} \cite{Grefenstette:2011:ESC:2145432.2145580,kartsadrqpl2014,milajevs-EtAl:2014:EMNLP2014}, \textbf{3K} \cite{Dinu:2010:MDS:1870658.1870771,milajevs-purver:2014:CVSC} or \textbf{10K} \cite{polajnar-clark:2014:EACL,Baroni2010nouns}. The most common reason thereof is that these models  assign tensors to functional words. For a vector space $V$ with $k$ dimensions,  a tensor $V \otimes V \cdots \otimes V$ of rank $n$ has $k^n$ dimensions. Adjectives and intransitive verbs have tensors of rank 2, transitive verbs are of rank 3; for coordinators, the rank can go up to 7. Taking $k = \textbf{200K}$ already results in a highly intractable tensor of $\textbf{8} \times \textbf{10}^{\textbf{15}}$ dimensions for a transitive verb.

An alternative way of obtaining a vector space with few dimensions, usually with just 100--500, is to use SVD as a part of Latent Semantic Analysis \cite{ARIS:ARIS1440380105} or another models such as SGNS \cite{mikolov2013efficient} and GloVe \cite{pennington2014glove}. However, these models take more time to instantiate in comparison to weighting of a co-occurrence matrix and bring more parameters to explore. In this work we focus on vector spaces that directly weight a co-occurrence matrix and report results for SVD, GloVe and SGNS from the study of \newcite{TACL570} for comparison.

The mismatch of experiments with non-dense models in vector dimensionality between lexical and compositional tasks gives rise to a number of questions:
\begin{compactitem}
\item To what extent does model performance depend on vector dimensionality?
\item Do parameters influence 200K and 1K dimensional models similarly? Can the findings of \newcite{TACL570} be directly applied to models with few thousand dimensions?
\item If not, can we derive suitable parameter selection heuristics which take account of dimensionality?
\end{compactitem}

@@

In addition to the extreme $1$ and $n$ weightings, we also experiment with a $\log n$ weighting.

@@

\newcite{TACL570} show that performance is affected by smoothing the context distribution $P(x)$:
%
\begin{equation}
  \label{eq:cds}
  P_{\alpha}(x) = \frac{n(x)^{\alpha}}{\sum_{c}n(c)^{\alpha}}
\end{equation}
%
We experiment with $\alpha$=1 (no smoothing) and $\alpha = 0.75$. We call this estimation method \emph{local context probability}; we can also estimate a \emph{global context probability} based on the size of the corpus $C$:
%
\begin{equation}
  \label{eq:cds-nan}
  P(x) = \frac{n(x)}{|C|}
\end{equation}

@@

We evaluate these heuristics by comparing the performance they give on SimLex-999 against that obtained using the best possible parameter selections (determined via an exhaustive search at each dimensionality setting). We also compare them to the best scores reported by \newcite{TACL570} for their model, word2vec-SGNS \cite{mikolov2013efficient} and GloVe \cite{pennington2014glove} -- see Figure~\ref{fig:best-simlex} (only the better-performing SPMI and SCPMI are shown).

@@

Finally, to see whether the heuristics transfer robustly, we repeat this comparison on the MEN dataset (see Figures~\ref{fig:best-men}, \ref{fig:global-best-men}). Again, PMI and CPMI follow the best possible setup, with SPMI and SCPMI showing only a slight drop below ideal performance; and again, the heuristic settings give performance close to the optimum, and significantly higher than average or standard parameters.

\begin{table}
  \centering
  \begin{tabular}{lll}
    \toprule
    Model           & SimLex-999 & MEN \\
    \midrule
    PPMI$^*$        & 0.393      & 0.745 \\
    SVD$^*$         & 0.432      & 0.778 \\
    SGNS$^*$        & 0.438      & 0.774 \\
    GloVe$^*$       & 0.398      & 0.729 \\
    \addlinespace
    This work       & 0.385      & 0.765 \\
    \bottomrule
  \end{tabular}
  \caption{\textbf{Results.} Our model in comparison to the previous work. On the similarity dataset our model is 0.008 points behind a PPMI model, however on the relatedness dataset 0.020 points above, note the difference in dimensionality and source corpora, and window size. SVD and SGNS and GloVe numbers are given for comparison. $^*$Results reported by \newcite{TACL570}.}
\label{tab:results}
\end{table}

@@

This paper presents a systematic study of co-occurrence quantification focusing on the selection of parameters presented in \newcite{TACL570}. We replicate their recommendation for high-dimensional vector spaces, and show that with appropriate parameter selection it is possible to achieve comparable performance with spaces of dimensionallity of 1K to 50K, and propose a set of model selection heuristics that maximizes performance. We foresee the results of the paper are generalisable to other experiments, since model selection was performed on a similarity dataset, and was additionally tested on a relatedness dataset.

In general, model performance depends on vector dimensionality (the best setup with 50K dimensions is better than the best setup with 1K dimensions by 0.03 on SimLex-999). Spaces with few thousand dimensions benefit from being dense and unsmoothed  ($k < 1$, global context probability); while high-dimensional spaces are better sparse and smooth  ($k > 1$, $\alpha = 0.75$). However, these heuristics do not guarantee the best possible result for unweighted and $n$-weighted models because of the high variance of the corresponding scores. Based on this we suggest the use of \logNSPMI/ or \logNSCPMI/ with dimensionality of at least 20K to ensure good performance on lexical tasks.
